The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2006-10-25 trunk/vcd/inst/doc/strucplot.Rnw
new strucplot vignette added
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@608 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\class}[1]{\textsf{#1}}
\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
\newcommand{\data}[1]{`\texttt{#1}'}

%% \code without `-' ligatures
\def\nohyphenation{\hyphenchar\font=-1 \aftergroup\restorehyphenation}
\def\restorehyphenation{\hyphenchar\font=`-}
{\catcode`\-=\active%
  \global\def\code{\bgroup%
    \catcode`\-=\active \let-\codedash%
    \Rd@code}}
\def\codedash{-\discretionary{}{}{}}
\def\Rd@code#1{\texttt{\nohyphenation#1}\egroup}

@@

@@

%\VignetteIndexEntry{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
%\VignetteDepends{vcd}
%\VignetteKeywords{graphics, categorical data}
%\VignettePackage{vcd}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
data(HairEyeColor)
@

\title{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
\author{by David Meyer, Achim Zeileis, and Kurt Hornik}
\date{}

\maketitle
\sloppy

\section{Framework Overview}

The strucplot framework in the {\sf R} package \pkg{vcd} visualizes
multi-way contingency tables and integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation that can be visualized in
ways similar to a two-dimensional table.
This principle defines a class of conditional displays which allows
for granular control of graphical appearance aspects, including:

@@

\noindent This document gives an introduction to the framework,
whereas labeling and shading issues are described in separate vignettes.

The strucplot framework is highly modularized: Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot. These are examples of `graphical appearance control'
(`grapcon') functions. They are created by generating functions, allowing
flexible parameterization and extensibility. 
The first part of their names (\code{\emph{group\_foo()}}) 
reflects the group they belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{\emph{struc\_foo()}}, 
\code{\emph{labeling\_foo()}}, and \code{\emph{legend\_foo()}}) 
directly produce graphical output (``add ink to the canvas''), whereas
the parameter functions
(created by \code{\emph{spacing\_foo()}} and \code{\emph{shading\_foo()}}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{\emph{struc\_foo()}} implement the core functionality, creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulas. Finally, on the fourth
level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

@@

\section{Mosaic, Association, and Sieve Plots}

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous variable (sex, i.e., gender). The
`flattened' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base {\sf R}, but allowing the specification of split directions):

<<structable,results=verbatim>>=
(hec <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
@

Let us first visualize the contingency table by means of a mosaic plot
\citep{vcd:Hartigan+Kleiner:1984} which is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. Figure \ref{fig:observed} depicts 
the effect of

<<Observed>>=
mosaic(hec)
@

@@

\begin{figure}[p]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE>>=
<<Observed>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data.}
\label{fig:observed}
\end{center}
\end{figure}

\noindent The small bullets indicate zero entries in the corresponding
cell. Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
{\sf R}, the default split direction and level ordering in all strucplot
displays correspond to the textual representation.
It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):

<<Expected>>=
mosaic(hec, type = "expected")
@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE>>=
<<Expected>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}

\noindent In order to compare observed and expected values,
a sieve plot \citep{vcd:riedwyl+schuepbach:1994} 
could be used (see Figure \ref{fig:sieve}):

<<sieve>>=
sieve(hec)
@

@@

\noindent Alternatively, we can directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
base {\sf R}, \pkg{vcd}'s \codefun{assoc}
function scales to more than two variables (see Figure \ref{fig:residuals}):

<<Residuals>>=
assoc(hec, compress = FALSE)
@

@@

\noindent The \code{compress} argument keeps distances between tiles
equal for better comparison.

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument (default: alternating splits starting
with a vertical one).

<<split1>>=
mosaic(hec, split_vertical = c(TRUE, FALSE, TRUE))
@

\noindent For compatibility with \codefun{mosaicplot} in base {\sf R},
the \codefun{mosaic} function also allows the use of a \code{"direction"} 
argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figure \ref{fig:split}):

<<split2>>=
mosaic(hec, direction = c("v","h","v"))
@

@@

By a suitable combination of splitting, spacing, and
labeling settings, the functions provided by the strucplot framework
can be customized in a quite flexible way. For example, 
\codefun{doubledecker} is simply a wrapper for 
\codefun{mosaic}, setting the right defaults. 
Figure \ref{fig:doubledecker} shows a doubledecker plot of
the \data{Titanic} data, explaining the probability of survival (`survived') by age, given
sex, given class. It is created by:

@@

\begin{figure}[h]
\begin{center}
<<doubledeckerfig,fig=TRUE,echo=FALSE>>=
<<doubledecker1>>
@
\caption{Doubledecker plot for the \data{Titanic} data.}
\label{fig:doubledecker}
\end{center}
\end{figure}

\section{Conditional and partial views}

So far, we have visualized full tables. For objects of class \class{table},
conditioning on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is simply done by indexing.
However, subsetting \code{"structable"} objects is more restrictive 
because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
conditioning is only possible ``outside-in'':

<<subsetting,results=verbatim>>=
hec
hec["Male",]
hec[c("Male", "Brown"),]
hec["Male","Green"]
@

@@

<<Variables>>=
pushViewport(viewport(layout = grid.layout(ncol = 2)))

pushViewport(viewport(layout.pos.col = 1))
mosaic(hec["Male"], margins = c(top = 2, 0), sub = "Male", newpage = FALSE)
popViewport()

pushViewport(viewport(layout.pos.col = 2))
mosaic(hec["Female"], margins = c(top = 2, 0), sub = "Female", newpage = FALSE)
popViewport(2)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=5,width=12>>=
<<Variables>>
@
\caption{Distribution of hair and eye color, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent Note the use of the \code{margins} argument: it takes a
vector with up to four values whose unnamed components are recycled,
but ``overruled'' by the named arguments. Thus, in the example, only
the top margin is set to 2 lines, and all other to 0.
This idea applies to almost all
vectorized arguments in the strucplot framework (with
\code{split\_vertical} as a prominent exception).

Since mosaic displays are ``conditional plots'' by definition, we can
also use one single mosaic for stratified plots.
The formula interface of \codefun{mosaic} allows the
specification of conditioning variables (see Figure \ref{fig:conditioning}):

<<Conditioning>>=
mosaic(~ Hair + Eye | Sex, data = hec, split_vertical = TRUE, keep_aspect_ratio = FALSE)
@

>>>>

\documentclass{Z}

\usepackage{thumbpdf}

%% almost as usual
\author{David Meyer\\Wirtschaftsuniversit\"at Wien \And 
        Achim Zeileis\\Wirtschaftsuniversit\"at Wien \And 
        Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
  Tables with \pkg{vcd}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{David Meyer, Achim Zeileis, Kurt Hornik} %% comma-separated
\Shorttitle{The Strucplot Framework} %% a short title (if necessary)
\Plaintitle{The Strucplot Framework: Visualizing Multi-way Contingency
  Tables with vcd}

%% an abstract and keywords
\Abstract{
This paper has been published in the Journal of Statistical Software
\citep{vcd:Meyer+Zeileis+Hornik:2006b} and 
describes the ``strucplot'' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The framework's modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of ``graphical
appearance control'' functions. 
The framework is provided by the \proglang{R} package \pkg{vcd}.
}

\Keywords{contingency tables, mosaic plots, association plots, sieve plots, 
categorical data, independence, conditional independence, HSV, HCL,
residual-based shading, \pkg{grid}, \proglang{R}}
\Plainkeywords{contingency tables, mosaic plots, association plots,
  sieve plots, categorical data, independence, 
  conditional independence, HSV, HCL, residual-based shading, grid, R}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
data(HairEyeColor)
data(PreSex)
data(Arthritis)
art <- xtabs(~Treatment + Improved, data = Arthritis)
@

\newcommand{\var}[1]{\textit{\texttt{#1}}}
\newcommand{\data}[1]{\texttt{#1}}
\newcommand{\class}[1]{\textsf{#1}}
%% \code without `-' ligatures
\def\nohyphenation{\hyphenchar\font=-1 \aftergroup\restorehyphenation}
\def\restorehyphenation{\hyphenchar\font=`-}
{\catcode`\-=\active%
  \global\def\code{\bgroup%
    \catcode`\-=\active \let-\codedash%
    \Rd@code}}
\def\codedash{-\discretionary{}{}{}}
\def\Rd@code#1{\texttt{\nohyphenation#1}\egroup}

@@

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\section[Introduction]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

In order to explain multi-dimensional categorical data, statisticians
typically look for (conditional) independence structures. Whether the
task is purely exploratory or model-based, techniques such as mosaic and
association plots offer good support for visualization.  Both visualize
aspects of (possibly higher-dimensional) contingency tables, with
several extensions introduced over the last two decades, and
implementations available in many statistical environments.  A
\emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically an
area-proportional visualization of (typically, observed) frequencies,
composed of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a rectangle. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. An \emph{association plot}
\citep{vcd:Cohen:1980} visualizes the standardized deviations of
observed frequencies from those expected under a certain independence
hypothesis.  Each cell is represented by a rectangle that has (signed)
height proportional to the residual and width proportional to the square
root of the expected counts, so that the area of the box is proportional
to the difference in observed and expected frequencies.

Extensions to these techniques have mainly focused on the following
aspects.

\begin{enumerate}
\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
  double-decker plots \citep{vcd:hofmann:2001}, spine plots, or
  spinograms \citep{vcd:hofmann+theus}.
\item Using residual-based shadings to visualize log-linear models
  \citep{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \citep{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models 
  \citep{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Meyer+Zeileis+Hornik:2006b}.
\end{enumerate}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \pkg{ViSta} \citep{vcd:young:1996}, 
\pkg{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \citep{vcd:Theus:2003},
\proglang{R} \citep{vcd:R:2006}, and \proglang{S-PLUS}
\citep{vcd:SPLUS:2005}. For \proglang{R}, currently three
implementations do exist in the packages \pkg{graphics} (in base \proglang{R}), 
\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006b},
and \pkg{iplots} \citep{vcd:urbanek+wichtrey:2006}, respectively.
Table \ref{tab:compare} gives an overview of the available functionality in these systems.
Most environments are available on Windows, MacOS, and Linux/Unix
variants, except \pkg{MANET} which is only available for the
Macinthosh platforms.

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|c|c|c|c|}
    \hline
                             &                &                   
                             &\multicolumn{3}{c|}{}   &        & &\\
                             & \proglang{SAS} & \proglang{S-PLUS} 
                             &\multicolumn{3}{c|}{\proglang{R}} & \pkg{ViSta} & \pkg{MANET}
                             & \pkg{Mondrian}\\
                             &                &                   
                             &\pkg{base}&\pkg{vcd}   &\pkg{iplots}&        & &\\\hline

    Basic functionality      &  $\times$  & $\times$  & $\times$ &$\times$ &$\times$ &   $\times$   & $\times$& $\times$\\
    Shape                    &            &           &          &$\times$ && $\times$  &  $\times$&\\
    Res.-based shadings  &  $\times$  &        & $\times$ & $\times$ & ($\times$) &     &($\times$)& ($\times$)\\
    Highlighting             &     &        &   &$\times$   &$\times$  & $\times$ & $\times$& $\times$\\
    Conditional views        &  $\times$  &        &    &$\times$ &          & $\times$ &  $\times$&\\
    Interaction              &     &        &   &  &$\times$  &   $\times$   & $\times$& $\times$\\
    Linking                  &     &        &   &  &$\times$  & $\times$ & $\times$& $\times$\\
    Extensible design        &     &        &  &$\times$ & &       &  &\\
    Language                 & \proglang{SAS} & \proglang{S} &
    \proglang{R} & \proglang{R} & \proglang{R}/\proglang{Java} & \proglang{XLisp} & \proglang{C++} & \proglang{Java}\\
    \hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions. Figure~\ref{fig:arthritis} shows the results from a
double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. The dark
blue cell indicates that the rate of treated patients showing marked improvement
is significant at the 1\% level. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null hypothesis) causes the rejection of the
hypothesis of conditional independence. Figure~\ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The $\chi^2$ test 
of independence, based on the permutation distribution,  
rejects the null hypothesis: possibly, because
the tendency of people to have extramarital
sex when they had premarital sex 
is particularly marked among married people? The rate of such women and men
ist significant at the 0.01 and 0.1 level, respectively.
Finally, Figure~\ref{fig:titanic} visualizes the ``Survival on the Titanic'' 
data using a double-decker plot. Here, a binary response (survival of the disaster)
is to be explained by other factors (class, gender, and age). The gray
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, gp = shading_max, split_vertical = TRUE)
@
\caption{Mosaic plot for the \data{Arthritis} data.}
\label{fig:arthritis}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(UCBAdmissions, panel = cotab_coindep, shade = TRUE, legend = FALSE, 
          type = "assoc")
@
\caption{Conditional association plot for the \data{UCBAdmissions} data.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
presextest <- coindep_test(PreSex, margin = c(1,4), 
                           indepfun = function(x) sum(x^2), n = 5000)
mosaic(PreSex, condvars = c(1, 4), shade = TRUE, 
       gp_args = list(p.value = presextest$p.value))
@
\caption{Mosaic plot for the \data{PreSex} data.}
\label{fig:presex}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<Titanic,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
doubledecker(Survived ~ ., data = Titanic, labeling_args = 
             list(set_varnames = c(Sex = "Gender")))
@
\caption{Double-decker plot for the \data{Titanic} data.}
\label{fig:titanic}
\end{center}
\end{figure}

This paper describes the strucplot framework provided by the \pkg{vcd}
package for the \proglang{R} environment for statistical computing and
graphics, available from the Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}).  The framework integrates displays
such as mosaic, association, and sieve plots by their unifying property
of being flat representations of contingency tables.  These basic plots,
as well as specialized displays for conditional independence, can be
used both for exploratory visualization and model-based analysis.
Exploratory techniques include specialized displays for the bivariate
case, as well as pairs and trellis-type displays for
higher-dimensional tables. Model-based tools include methods suitable
for the visualization of conditional independence tests (including
permutation tests), as well as for the visualization of particular GLMs
(logistic regression, log-linear models). Additionally, two of the
framework's further strengths are its flexibility and extensibility:
graphical appearance aspects such as shading, labeling, and spacing are
modularized by means of ``\underline{\vphantom{g}gr}aphical 
\underline{\vphantom{g}ap}pearance \underline{\vphantom{g}con}trol''
(\emph{grapcon}) functions, allowing
fine-granular customization and user-level extensions.

The remainder of the paper is organized as follows. In
Section \ref{sec:strucplot}, we give an overview of the strucplot
framework, describing the hierarchy of the main components and the
basic functionality. 
In Section \ref{sec:shading}, we demonstrate how
(residual-based) shadings support the visualization of log-linear models and
the results of independence tests. Also, we explain step-by-step how
the concepts of generating and grapcon functions
can be combined to provide a flexible customization of
complex graphical displays as created by the strucplot framework.
Sections \ref{sec:labeling} and \ref{sec:spacing} discuss in detail the labeling
and spacing features, respectively. Section \ref{sec:example}
exemplifies the framework in the analysis of a four-way data set.
Section \ref{sec:conclusion} concludes the work.

\section[The strucplot framework]{The strucplot framework}
\label{sec:strucplot}

The strucplot framework in the \proglang{R} package \pkg{vcd}, used for visualizing
multi-way contingency tables, integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into rows and columns using recursive
conditional splits, given the margins. The result is a 
``flat'' representation that can be visualized in
ways similar to a two-dimensional table.
This principle defines a class of conditional displays which allows
for granular control of graphical appearance aspects, including:

@@

The strucplot framework is highly modularized: Figure~\ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot (see Table \ref{tab:grapcons} for an overview). 
These are examples of grapcon functions. They are created by generating functions
(\emph{grapcon generators}), allowing
flexible parameterization and extensibility (Figure~\ref{fig:struc}
only shows the generators). The generator names
follow the naming convention \code{\textit{group\_foo}()},  
where \code{\textit{group}} reflects the group the
generators belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\textit{foo}()}, 
\code{labeling\_\textit{foo}()}, and \code{legend\_\textit{foo}()}) 
directly produce graphical output (i.e., ``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\textit{foo}()} and \code{shading\_\textit{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\textit{foo}()} implement the core functionality, 
creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure~\ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulae. Finally, on the fourth
level, there are ``related'' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

\begin{table}
  \begin{tabular}{|l|l|l|}
    \hline
    \textbf{Group} & \textbf{Grapcon generator} & \textbf{Description}\\\hline
    strucplot & \codefun{struc\_assoc} & core function for association plots\\
    core      & \codefun{struc\_mosaic} & core function for mosaic plots\\
              & \codefun{struc\_sieve} & core function for sieve plots\\\hline\hline
     labeling & \codefun{labeling\_border} & border labels\\
              & \codefun{labeling\_cboxed} & centered labels with
              boxes, all labels clipped,\\
              && and on top and left border\\
              & \codefun{labeling\_cells} & cell labels\\
              & \codefun{labeling\_conditional} & border labels 
                                                  for conditioning variables\\
              && and cell labels for conditioned variables\\
              & \codefun{labeling\_doubledecker} & draws labels for
              doubledecker plot\\
              & \codefun{labeling\_lboxed} & left-aligned labels with boxes\\
              & \codefun{labeling\_left} & left-aligned border labels\\
              & \codefun{labeling\_left2} & left-aligned border
              labels, all labels on top and left border\\
              & \codefun{labeling\_list} & draws a list of labels
              under the plot\\\hline\hline
     shading  & \codefun{shading\_binary} & visualizes the sign of the  residuals\\
              & \codefun{shading\_Friendly} & implements Friendly
              shading (based on HSV colors)\\
              & \codefun{shading\_hcl} & shading based on HCL colors\\
              & \codefun{shading\_hsv} & shading based on HSV colors\\
              & \codefun{shading\_max} & shading visualizing the
              maximum test statistic\\ && (based on HCL colors)\\
              & \codefun{shading\_sieve} & implements Friendly
              shading customized for sieve plots\\ && (based on HCL colors)\\\hline\hline
     spacing  & \codefun{spacing\_conditional} & increasing spacing
              for conditioning variables,\\&& equal spacing for
              conditioned variables\\
              & \codefun{spacing\_dimequal} & equal spacing for each dimension\\
              & \codefun{spacing\_equal} & equal spacing for all dimensions\\
              & \codefun{spacing\_highlighting} & increasing spacing,
              last dimension set to zero\\
              & \codefun{spacing\_increase} & increasing spacing\\\hline\hline
     legend   & \codefun{legend\_fixed} & creates a fixed number of
              bins (similar to \codefun{mosaicplot})\\
              & \codefun{legend\_resbased} & suitable for an
              arbitrary number of bins\\&& (also for continuous shadings)\\\hline

  \end{tabular}
  \caption{Available grapcon generators in the strucplot framework}
  \label{tab:grapcons}
\end{table}

@@

\subsection{Mosaic, association, and sieve plots}

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous gender variable (\code{Sex}). The
``flattened'' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base \proglang{R}, but allowing the specification of split directions):

<<structable,echo=TRUE,results=verbatim>>=
(HEC <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
@

Let us first visualize the contingency table by means of a mosaic plot.
% \citep{vcd:Hartigan+Kleiner:1984} which is basically 
% an area-proportional visualization of (typically, observed) frequencies, composed
% of tiles (corresponding to the cells) created by recursive
% vertical and horizontal splits of a square. Thus, the area of each tile
% is proportional to the corresponding cell entry \emph{given} the
% dimensions of previous splits. 
The effect of

<<Observed>>=
mosaic(HEC)
@

@@

%\setkeys{Gin}{width=0.75\textwidth}
\begin{figure}[p]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE,height=6>>=
<<Observed>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data.}
\label{fig:observed}
\end{center}
\end{figure}

\noindent depicts the observed frequencies of the \code{HairEyeColor}
data. If there are zero entries, tiles have zero area and are, additionally,
marked by small bullets (see, e.g, Figure~\ref{fig:titanic}). By
default, these cells are not split further. The bullets help
distinguishing very small cells from zero entries, and are
particularly useful when color shadings come into play 
(see the example using the \data{Bundesliga} data in Section \ref{sec:overview}). 
Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
\proglang{R}, the default split direction and level ordering in all strucplot
displays correspond to the textual representation produced by the
print methods.
It is also possible to visualize the expected values instead of the
observed values (see Figure~\ref{fig:expected}):

<<Expected>>=
mosaic(HEC, type = "expected")
@

\begin{figure}[p]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE,height=6>>=
<<Expected>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}
%\setkeys{Gin}{width=0.7\textwidth}

\noindent In order to compare observed and expected values,
a sieve plot \citep{vcd:riedwyl+schuepbach:1994} 
could be used (see Figure~\ref{fig:sieve}):

<<sieve>>=
sieve(~ Sex + Eye + Hair, data = HEC, spacing = spacing_dimequal(c(2,0,0)))
@

@@

\noindent where \code{spacing\_dimequal} is used to set the spacing
of the second and third dimension to zero. Alternatively, we can directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are conveniently visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
base \proglang{R}, \pkg{vcd}'s \codefun{assoc}
function scales to more than two variables (see Figure~\ref{fig:residuals}):

<<Residuals>>=
assoc(HEC, compress = FALSE)
@

@@

\noindent where the \code{compress} argument keeps distances between tiles
equal.

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument. The default is to alternate splits starting
with a horizontal one (see Figure~\ref{fig:split}):

<<echo=FALSE>>=
options(width=60)
@

<<split1>>=
mosaic(HEC, split_vertical = c(TRUE, FALSE, TRUE), 
labeling_args = list(abbreviate = c(Eye = 3)))
@

<<echo=FALSE>>=
options(width=70)
@

@@

\noindent (Note that \code{HEC}, a \class{structable} object, already
includes a splitting information which simply gets overloaded in this
example.)  For compatibility with \codefun{mosaicplot} in base
\proglang{R}, the \codefun{mosaic} function also allows the use of a
\code{direction} argument taking a vector of \code{"h"} and \code{"v"}
characters:

<<split2>>=
mosaic(HEC, direction = c("v","h","v"))
@

By a suitable combination of splitting, spacing, and
labeling settings, the functions provided by the strucplot framework
can be customized in a quite flexible way. For example, 
the default method for \codefun{doubledecker} is simply a wrapper for 
\codefun{strucplot}, setting the right defaults.
Most default settings such as colors, spacing, and labeling are
specified via the parameters and passed through to
\codefun{strucplot}. The additional code just handles the dependent
variable information, and in particular permutes the table to have the dependent
variable as the last dimension as required for the doubledecker plot.
Figure~\ref{fig:titanic} shows a doubledecker plot of
the \data{Titanic} data, explaining the probability of survival (``survived'') by age, given
sex, given class. It is created by:

@@

\subsection{Conditional and partial views}

So far, we have visualized either full or collapsed tables, as
suggested by the analysis task at hand. Subtables can be selected in a
similar way as for objects of class \class{table} using
indexing. Note, however, that subsetting of \class{structable} objects 
is more restrictive because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
subsetting is only possible ``outside-in'', that is, indexing operates
on blocks defined by the variable levels. In the following, we use the
Titanic data again, this time collapsed over \code{Survived} to investigate the
structure of crew and passengers (and having the \code{Child} and
\code{Age} labels of the \code{Age} variable swapped for optical clarity):

<<echo=FALSE>>=
options(width=75)
@

<<subsetting,results=verbatim>>=
(STD <- structable(~ Sex + Class + Age, data = Titanic[,,2:1,]))
STD["Male",]
STD["Male", c("1st","2nd","3rd")]
@

<<echo=FALSE>>=
options(width=70)
@

\noindent \emph{Conditioning} on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is done using the \code{[[} operator. %]]
Here again, the sequence of conditioning levels is restricted by the
hierarchical structure of the \class{structable} object. In the
following examples, note that compared to subsetting, the first
dimension(s) are dropped:

<<conditioning,results=verbatim>>=
STD[["Male",]]
STD[[c("Male", "Adult"),]]
STD[["Male","1st"]]
@

@@

<<Variables1>>=
pushViewport(viewport(layout = grid.layout(ncol = 2)))
@

<<Variables2>>=
pushViewport(viewport(layout.pos.col = 1))
mosaic(STD[["Male"]], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
popViewport()
@

<<Variables3>>=
pushViewport(viewport(layout.pos.col = 2))
mosaic(STD[["Female"]], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
popViewport(2)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
<<Variables1>>
<<Variables2>>
<<Variables3>>
@
\caption{Two mosaic displays put side-by-side, visualizing the 
distribution of class and age, given gender. The marginal
distribution of gender cannot be seen.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent Note the use of the \code{margins} argument: it takes a
vector with up to four values whose unnamed components are recycled,
but ``overruled'' by the named arguments. Thus, in the second example, only
the top margin is set to 2.5 lines, and all other to 0.
This idea applies to almost all
vectorized arguments in the strucplot framework (with
\code{split\_vertical} as a prominent exception).

The \codefun{cotabplot} function does a much better job on this task:
it arranges stratified strucplot displays 
in a lattice-like layout, conditioning on variable \emph{levels}.
The plot in Figure~\ref{fig:cotabplot} shows hair and eye color, given sex:
